Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-01-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.
Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. Elevated CO~2~ may benefit understory seedlings operating close to their carbon compensation point, but given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

@@

Leaf hydraulic, stomatal conductance, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light, and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, when water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves typically at noon, midday stomatal depression is more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
In sun leaves, $T_{leaf}$ thus further increases due to the lack of transpirational cooling [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is maximal, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a].

Further, tree height also plays a role in determining the gradient of $g_s$ vertically, which has implications for $T_{leaf}$. With an increase in tree height, water travels a longer distance to meet the evaporative demands in sunlit canopy leaves. Longer water path length, higher irradiance and VPD imposes hydraulic constraints in sunlit canopy leaves [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Therefore, although typically $g_s$ increases with light, upper canopy leaves of taller trees have lower $g_s$ and leaf water potential to enable high leaf water-use efficiency [@kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a]. This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of canopy leaves more than understory leaves, particularly when water availability is limited.

>>>>

Atmospheric concentrations of carbon dioxide (CO~2~) and water vapor can also vary across the vertical gradient.
Atmospheric CO~2~ concentrations tend to be higher near the ground at night, associated with plant and soil respiration, although the elevated CO~2~ quickly dissipates during the day [@koike_leaf_2001; @yang_spatial_1999]. Vertical CO~2~ concentration gradients can influence the ratio of leaf inter-cellular CO~2~ to ambient CO~2~ ($c_{i}/c_{a}$), and refixation of respired CO~2~ inferred by $\delta^{13}C$, where both tend to be higher in the lower canopy positions [Table 1]. Particularly, understory seedlings above ground benefit from elevated CO~2~ by having highest rates of CO~2~ refixation [Brooks et al. 1997]. However, given that differences are negligible for most of the day when photosynthesis is occurring, gradients in CO~2~ concentrations are unlikely to significantly affect the energy balance and metabolism of leaves across the forest vertical gradient.

@@

Leaf hydraulic, stomatal conductance, and $g_b$ are all critical for latent heat loss, which in turn can be important for regulating $T_{leaf}$ (Fig. `r fig_leaf_T`). 
Due to higher wind, lower RH, and smaller leaf sizes in the canopy (Fig. `r fig_NEON_vertical`, Table `r table_leaf_traits`), $g_b$ increases with height (Table `r table_leaf_metabolism`).
Maximum $g_s$ increases with light--– typically tracking photosynthetic capacity (Wong et al. 1979)-– and is thus higher in the sun-exposed upper canopy than in sub-canopy or understory leaves (Table `r table_leaf_metabolism`). 
However, water supply cannot meet the demands incurred by the high irradiance, VPD and $g_b$ experienced by sun leaves with fully open stomata because of height-related constraints on water transport [ **e.g, Yoder et al. 1994, Koch et al. 2004, Sillet et al. 2010**] and/or because of ncreased evaporative demand ($VPD_{leaf}$) caused by leaf warming in sunlit canopy locations. These constraints tend to educe leaf water potential making midday stomatal depression more prevalent in sun leaves than shade leaves in closed-canopy forests (Table `r table_leaf_metabolism`), which drives the lower intracellular CO~2~ and carbon isotope discrimination discussed in section 3.1 (Table 1).
Stomatal depression duces transpirational cooling, thus amplifying the warming of sun leaves by high radiation loads [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994; @kosugiSeasonalFluctuationsTemperature2006]. 
The temperature at which $g_s$ is greatest, $T_{opt}$ of $g_s$, did not differ significantly between sun and shade leaves in three tropical tree species [@slotPhotosyntheticHeatTolerance2019], but sun leaves in the upper canopy show a stronger decrease in $g_s$ in response to rising $T_{air}$ (driven y increased VPD) than do shade leaves in the lower canopy [@hernandezSimilarTemperatureDependence2020; @carterExperimentalWarmingTropical2021a]. This, added to the tendency for sun leaves to have higher $T_{leaf}$, implies that high $T_{air}$ should decrease $g_s$ of upper canopy leaves more than understory leaves, particularly when water availability is limited.
